---
title: "US coal exports"
author: "Joel Bazzle"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: yeti
    highlight: haddock 
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---

```{r setup, include=FALSE}
## These next lines set the default behaviour for all R chunks in the .Rmd document.
## I recomend you take a look here: https://rmarkdown.rstudio.com/authoring_rcodechunks.html
knitr::opts_chunk$set(
  echo = TRUE, ## Show all R output
  cache = TRUE, ## Cache the results to increase performance.
  message = FALSE ## Suppress messages (e.g. about namespace conflicts)
  ) 
knitr::opts_chunk$set(cache=TRUE) 
```

## Preliminaries: 

### Load libraries

It's a good idea to load your libraries at the top of the Rmd document so that everyone can see what you're using. Similarly, it's good practice to set `cache=FALSE` to ensure that the libraries are dynamically loaded each time you knit the document.

```{r libs, cache=FALSE}
## Install the pacman package if necessary
if (!require("pacman")) install.packages("pacman")
## Install other packages using pacman::p_load()
pacman::p_load(httr, readxl, here, tidyverse, data.table, janitor, dplyr, viridis, ggthemes, RColorBrewer, gganimate, gifski, png)
```

### Read in the data

Use `httr::GET()` to fetch the EIA excel file for us from web. (We'll learn more about `httr`, GET and other HTTP methods when we get to webscraping next week.) 

```{r get}
# library(here) ## Already loaded
# library(httr) ## Already loaded
url = "https://www.eia.gov/coal/archive/coal_historical_exports.xlsx"
## Only download the file if we need to
if(!file.exists(here::here("data/coal.xlsx"))) {
  GET(url, write_disk(here::here("data/coal.xlsx")))
}
```

Next, we read in the file.

```{r read}
# library(readxl) Already loaded
coal = read_excel(here::here("data/coal.xlsx"), skip=3, na=".")
```

We are now ready to go.

## 1) Clean the column names

The column (i.e. variable) names aren't great: Spacing, uppercase letters, etc. 
```{r names, dependson=coal}

coal = clean_names(coal)
  
names(coal)
```


## 2) Total US coal exports over time (year only)

Plot the US's total coal exports over time by year ONLY. What secular trends do you notice in the data?


```{r}
unique(coal$coal_origin_country)

## This is all export data, so that's cool

##multiple observations per year/total is region total
##note to future self: aggregate doesn't play nice with NA littered data, apparently

coalyeartotals = coal %>% 
  group_by(year) %>% 
  summarize(sum(total, na.rm = T))

colnames(coalyeartotals) = c("year", "total")

coalyeartotals %>% 
  ggplot(mapping = aes(x= year, y = total)) +
  geom_point(color = "blue")+
  geom_line(color = "red")+
  scale_y_continuous(labels = scales::comma)
```

**At first glance I thought there was a political trend because of the sudden rise in 2016, but that doesn't necessarily account for the increase between 2008-2012 and the fall between 2012-2016.  Attempts to find anything about domestic coal use from this period seem to be an excellent way to generate ad revenue for Politico whilst wading into their substance-free editorials written by right-wing hacks.**

**That being said, it seems like global demand has been steadily declining in most countries since 2012.  The largest outlier in that trend is China, which has significantly increased its coal use.  Coal exporters have struggled to increase exports to China because Oregon and Washington have blocked the construction of new export terminals due to obvious environmental concerns that would accompany both the terminal and the rail transport of coal.**

**In addition to the above, a bit more digging (and finally locating something from Reuters, which is thankfully still largely free from the influence of right-wing hack editorials) indicates that coal prices in Asia fell significantly between 2011-2016 and rose again between 2016-2018, which puts the graph into a sensible context.  Given India, Japan, and China are the countries buying the most US coal, such a price drop likely means that US coal was no longer able to compete in the region during this time frame, resulting in a significant fall in exports of US coal.**



## 3) Total US coal exports over time (year AND quarter)

Now do the same as the above, except aggregated quarter of year (2001Q1, 2002Q2, etc.). Do you notice any seasonality that was masked from the yearly averages?

```{r}

coalquartertotals = coal %>% 
  group_by(quarter, year) %>% 
  summarize(sum(total, na.rm = T))

colnames(coalquartertotals) = c("quarter", "year", "total")

coalquartertotals %>% 
  ggplot() +
  geom_point(aes(x = year, y = total, color = quarter, size = total))+
  geom_line(aes(x = year, y = total, color = quarter))+
  scale_y_continuous(labels = scales::comma)+
  scale_color_viridis(option = "D", limits = c(1, 4.5))+
  facet_wrap(~ quarter)

## Note for future self: coerce these god awful, incoherent, colorblind unfriendly pallets using limits = (interval)

```

*Hint: ggplot2 is going to want you to convert your quarterly data into actual date format before it plots nicely. (i.e. Don't leave it as a string.)*

**It looks like quarter 2 2012 was the actual export peak, and the fall began in quarter 3 2012.  Additionally, it looks like the most coal is generally exported in quarter 2**

## 4) Exports by destination country

### 4.1) Create a new data frame

Create a new data frame called `coal_country` that aggregates total exports by destination country (and quarter of year). Make sure you print the resulting data frame so that it appears in the knitted R markdown document.

```{r}

coal_country = coal %>% 
  group_by(year, quarter, coal_destination_country) %>% 
   summarize(sum(total, na.rm = T))

colnames(coal_country) = c("year", "quarter", "destination", "total")

coal_country

```

### 4.2) Inspect the data frame

It looks like some countries are missing data for a number of years and periods (e.g. Albania). Confirm that this is the case. What do you think is happening here?

```{r}
frequencytable = coal_country %>% plyr::count("destination")
frequencytable
```

**Based only on the first page output (conveniently containing Albania), it is clear that the frequency with which countries appear in the data varies significantly.  A reasonable explanation, I would think, is that exporters are not consistently exporting to every country on the list during every quarter of every year.  Albania, for instance, only imported US coal once**

### 4.3) Complete the data frame

Fill in the implicit missing values, so that each country has a representative row for every year-quarter time period. In other words, you should modify the data frame so that every destination country has row entries for all possible year-quarter combinations (from 2002Q1 through the most recent quarter). Order your updated data frame by country, year and, quarter. 


```{r}
coal_country = coal_country %>%  ungroup() %>% 
  complete(quarter, year, destination, fill = list(total = 0))

coal_country

```

**make sure that worked...**

```{r}
frequencytable2 = coal_country %>% plyr::count("destination")
frequencytable2
```


### 4.4 Some more tidying up

In answering the previous question, you _may_ encounter a situation where the data frame contains a quarter --- probably `r gsub("\\.", "q", lubridate::quarter(Sys.Date()-months(3), with_year = TRUE))` --- that is missing total export numbers for *all* countries. Did this happen to you? Filter out the completely missing quarter if so. Also: Why do you think this might have happened? (Please answer the latter question even if it didn't happen to you.) 

```{r}

gsub("\\.", "q", lubridate::quarter(Sys.Date()-months(3), with_year = TRUE))

```

**Complete() is finding all possible combinations of quarter, year, and destination.  We don't have data for 2020q4 in the data set, but 2020q4 is a possible combination, so when the complete function ran it generated rows for it.**

```{r}
coal_country = coal_country %>% 
  filter(!(year == 2020 & quarter == 4))

coal_country

```

**Looks like it is fixed now**

### 4.5) Culmulative top 10 US coal export destinations

Produce a vector --- call it `coal10_culm` --- of the top 10 top coal destinations over the full min(coal$year, na.rm=T)--\r max(coal$year, na.rm=T))` study period. What are they?

```{r}
coal_cum = coal_country %>% 
  group_by(destination) %>% 
   summarize(sum(total, na.rm = T))

colnames(coal_cum) = c("destination", "total")

coal_cum = coal_cum[order(-coal_cum$total),]

coal_culm = coal_cum[1:10, 1]

coal_culm

```
**The country list above represents the top ten buyers of US coals over the entire period.**

### 4.6) Recent top 10 US coal export destinations

Now do the same, except for most recent period on record (i.e. final quarter in the dataset). Call this vector `coal10_recent` and make sure to print it so that I can see it too. Are there any interesting differences between the two vectors? Apart from any secular trends, what else might explain these differences?

```{r}
coal_recent = coal_country %>% 
  filter((year == 2020 & quarter == 3))


coal10_recent = coal_recent[order(-coal_recent$total),]

coal10_recent = coal10_recent[1:10, 3]

coal10_recent

```


**Many of the differences are not surprising (Germany and the UK not being in the top ten, India being at the top).  Ukraine being in the top ten is pretty interesting.  I imagine they've shifted a fair portion of their energy dependence to the US in light of their closest trading partner launching an invasion and annexing part of their territory.  I also would have thought that Japan and South Korea would have been on a similar trajectory to the UK and Germany in terms of coal use, but this serves as evidence that assumption may not be correct.**


### 4.7) US coal exports over time by country

Plot the quarterly coal exports over time, but now disaggregated by country. In particular, highlight the top 10 (cumulative) export destinations and then sum the remaining countries into a combined "Other" category. (In other words, your figure should contain the time series of eleven different countries/categories.)

```{r}

coalbycountry = coal_country

coalbycountry$destination = ifelse(coalbycountry$destination %in% coal_culm$destination, coalbycountry$destination, "Other")

coalbycountry = coalbycountry %>% 
  group_by(year, destination) %>% 
  summarize(sum(total, na.rm = T))

colnames(coalbycountry) = c("year", "destination", "total")

coalbycountry %>% 
  ggplot() +
  geom_point(aes(x = year, y = total, color = destination))+
  geom_line(aes(x = year, y = total, color = destination))+
  scale_y_continuous(labels = scales::comma)+
  scale_color_viridis(discrete = T, option = "D")

```



**As someone that loves pasta, let me be the first to say that this spaghetti looks disgusting**

### 4.8) Make it pretty

Take your previous plot and add some swag to it. That is, try to make it as visually appealing as possible without overloading it with chart junk.

```{r}

coalbycountry$destination = factor(coalbycountry$destination, levels = c("Other", "Canada", "Netherlands", "Brazil", "India", "South Korea (Republic of Korea)", "United Kingdom", "Japan", "Italy", "Germany, Federal Republic of", "Mexico"))

coalbycountry %>% 
  ggplot(aes(x = year, y = total, fill = destination)) +
  geom_area()+
  scale_y_continuous(labels = scales::comma)+
  scale_fill_viridis(discrete = T, option = "cividis")+
  ggtitle("Top Importers of US Coal", subtitle= "2002-Present")+
  xlab("Year")+
  ylab("Tons of Coal")+
  theme_igray()
```

*Hint: You've got loads of options here. If you haven't already done so, consider a more bespoke theme with the `ggthemes`, `hrbrthemes`, or `cowplot` packages. Try out `scale_fill_brewer()` and `scale_colour_brewer()` for a range of interesting colour palettes. Try some transparency effects with `alpha`. Give your axis labels more refined names with the `labs()` layer in ggplot2. While you're at it, you might want to scale (i.e. normalise) your y-variable to get rid of all those zeros. You can shorten any country names to their ISO abbreviation; see `?countrycode::countrycode`. More substantively --- but more complicated --- you might want to re-order your legend (and the plot itself) according to the relative importance of the destination countries. See `?forcats::fct_reorder` or forcats::fct_relevel`.*

**To me, a stacked-area representation is a lot more coherent.  Lasagna does tend to be a bit more appealing than spaghetti, after all**

### 4.9) Make it interactive

Create an interactive version of your previous figure.

*Hint: Take a look at plotly::ggplotly(), or the gganimate package.*

```{r}
barplot = coalbycountry %>% 
  ggplot(aes(x = destination, y = total, fill = destination)) +
  geom_bar(stat = 'identity')+
  scale_y_continuous(labels = scales::comma)+
  scale_fill_viridis(discrete = T, option = "cividis")+
  theme_minimal()+
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+
  labs(title = 'year:{frame_time}', x = 'Country', y = 'Tons of Coal')+
  transition_time(year)+
  ease_aes('linear')

animate(barplot)


```


## 5) Show me something interesting

There's a lot still to explore with this data set. Your final task is to show me something interesting. Drill down into the data and explain what's driving the secular trends that we have observed above. Or highlight interesting seasonality within a particular country. Or go back to the original `coal` data frame and look at exports by customs district, or by coal type. Do we changes or trends there? Etcetera. Etcetera. My only requirement is that you show your work and tell me what you have found.

```{r}

ukrainedata = coal_country %>% 
  filter(destination == "Ukraine")

ukrainedata = ukrainedata %>% 
  group_by(year) %>% 
  summarize(sum(total, na.rm = T))

colnames(ukrainedata) = c("year", "total")

ukrainedata %>% 
  ggplot() +
  geom_point(aes(x = year, y = total))+
  geom_line(aes(x = year, y = total))+
  scale_y_continuous(labels = scales::comma)

```

** This doesn't exactly correspond to what I would have expected, so I dug into the tainted well of the internet again (while avoiding the vacuous black abyss that is Politico after what happened earlier in this assignment).  It turns out that Ukraine first started using coal-fired power plants in 2011 and then signed an agreement with China shortly thereafter.  Ukraine's domestic coal production fell significantly between 2014-2015 because a large portion of their production takes place in the occupied territory on the east side of the country.  Apparently they originally had to turn to Russia, of all places, to make up the deficit (after all, those Russian military operatives were just on vacation in eastern Ukraine and were in no way on an officially sanctioned mission).  In 2015-2016, they began importing much more coal from other sources, including the US.  I'm guessing the massive "falloff" for 2020 is simply due to the missing Q4 data. **
